Speeding up eQTL scans in the BxD population using GPU

### 2019-07-25 paper.tex
Added to the intro.

\author[$\ast$]{Xiaoqi Hu}
\author[$\ast$]{Hyeonju Kim}
\author[$\ast$]{Gregory Farage}
\author[$\dagger$]{Pjotr Prins}
\author[$\dagger$]{Robert W. Williams}
\author[$\S$]{Karl W. Broman}
\author[$\ast$,1]{Saunak Sen}

@@

\runningtitle{Speeding up eQTL scans in the BxD population using GPU} % For use in the footer

@@

\begin{abstract}
The BXD strains are an important reference population of
recombinant inbred lines which have been phenotyped extensively.
To facilitate interactive use of genotype-phenotype relationships
in this population, we sought to speed up eQTL scans where we
perform a univariate genome scan for every trait in a collection
of omic traits.  By using easily parallelizable operations such as
matrix multiplication, vectorized operations, and elementwise
operations, we are able to decrease runtimes approaching real-time
computation.  We used parallelization using different CPU threads
as well as GPUs.  We found that the speed advantage of GPUs is
dependent on problem size and shape.  These results indicate a pathway for
speeding up eQTL scans using LMMs.  Our implementation is in the
Julia programming language.

\end{abstract}

@@

\noindent        Computational demands in omics are growing exponentially. For
instance, eQTL analysis requires processing genomewide
expression and marker data; this can take minutes, hours or
days depending on problem size.  Scientists want to intereact
with the results and real-time processing would be valuable.
Our goal is to build a backend for GeneNetwork that will enable
real-time genome scans for important genetic reference
populations such as the BXD population.

>>>>

\author[$\ast$]{Xiaoqi Hu}
\author[$\ast$]{Hyeonju Kim}
\author[$\ast$]{Gregory Farage}
\author[$\dagger$]{Pjotr Prins}
\author[$\dagger$]{Robert W. Williams}
\author[$\S$]{Karl W. Broman}
\author[$\ast$,1]{\'Saunak Sen}

@@

\runningtitle{Speeding up eQTL scans in the BXD population using GPU} % For use in the footer

@@

\begin{abstract}
The BXD strains are an important reference population of
recombinant inbred lines which have been phenotyped extensively.
To facilitate interactive use of genotype-phenotype relationships
in this population, we sought to speed up eQTL scans where we
perform a univariate genome scan for every trait in a collection
of omic traits.  By using easily parallelizable operations such as
matrix multiplication, vectorized operations, and elementwise
operations, we are able to decrease runtimes approaching real-time
computation.  We used parallelization using different CPU threads
as well as GPUs.  We found that the speed advantage of GPUs is
dependent on problem size and shape.  These results indicate a pathway for
speeding up eQTL scans using LMMs.  Our implementation is in the
Julia programming language.
\end{abstract}

@@

\noindent The BXD mouse strains are an important set of recombinant
inbred lines with deep phenotypic characterization.  Since the strains
have been genotyped and sequenced, once a set of strains is
phenotyped, the data can be immediately used for QTL (quantitative
trait locus) mapping, and associational analysis with
previously-collected phenotypes.  For ``omic'' phenotypes collected
using high-throughput technologies additional analyses (such as
transcriptional network construction or causal mediation analyses) are
possible.  The GeneNetwork project ({\tt http://gn2.genenetwork.org})
seeks to faculitate this process by providing a searchable database of
phenotypes and genotypes in a variety of organisms (including mouse,
rat, and Arabidopsis).  It provides a suite of interactive tools for
browsing the data, QTL mapping, correlational analyses, network
construction, and genome browsing.  Our goal is to develop a real-time
backend for GeneNetwork to perform eQTL analysis (where we perform
single-trait QTL mapping for thousands of ``omic'' traits) for
important genetic reference populations such as the BXD population.

%% Computational demands in omics are growing exponentially. For
%% instance, eQTL analysis requires processing genomewide expression and
%% marker data; this can take minutes, hours or days depending on problem
%% size.  Scientists want to intereact with the results and real-time
%% processing would be valuable.  Our goal is to build a backend for
%% GeneNetwork that will enable real-time genome scans for important
%% genetic reference populations such as the BXD population.
